Embedding Intuitionistic into Classical Logic

### 2022-07-16 CSL/lipics-v2021-EIICL.tex
some more typos in the proofs

Suppose $\mathcal M, v\models\chi$. Then, for all $m\in M$ we have that $\mathcal M, v[m/a]\models\varphi[a/x]$ and therefore $\mathcal M_{\chi}, v[m/a]\models\varphi[a/x]^S_{Z\cup\{a\}}$ and thus $\mathcal M_{\chi}, v\models \forall x\varphi([a/x]^S_{Z\cup\{a\}}[x/a])$, i.e. $\mathcal M_\chi,v\models \chi^S_Z$. On the other hand suppose $\mathcal M, v\not\models\chi$. Then there exists $m\in M$ such that $\mathcal M, v[m/a]\not\models\varphi[a/x]$ and therefore $\mathcal M_\chi, v[m/a]\not\models\varphi[a/x]^H_{Z\cup\{a\}}$ and so by definition $\mathcal M_\chi, v\not\models\varphi[s(t_1,\dots t_n)/x]^H_Z$, i.e. $\mathcal M_\chi, v\not\models(\forall x\varphi)^H_Z$.

@@

\begin{definition}\label{def:transf-structure}
	For every classical and intuitionistic structure $\mathcal M$ define a structure $\mathcal S(\mathcal M,\varphi)$ that agrees with $\mathcal M$ on everything except the interpretation(s) of atoms of the form $P_\psi$. By slight abuse of notation we denote with $\vec z_\psi$ elements of the domain instead of variables. In the classical case have
	$$\begin{matrix}
		P_A^I(\vec z_A):\Leftrightarrow A^I(\vec z_A)\\
		P_{\varphi\wedge\psi}^I(\vec z_{\varphi\wedge\psi}) :\Leftrightarrow P_{\varphi}^I(\vec z_\varphi)\wedge P_{\psi}^I(\vec z_\psi)\indent P_{\varphi\vee\psi}^I(\vec z_{\varphi\vee\psi}) :\Leftrightarrow P_{\varphi}^I(\vec z_\varphi)\vee P_{\psi}^I((\vec z_\psi))\\P_{\varphi\to\psi}^I(\vec z_{\varphi\to\psi}) :\Leftrightarrow (\neg P_{\varphi}^I(\vec z_{\varphi}))\vee P_{\psi}^I(\vec z_{\psi})\\P_{\forall x\varphi}^I(\vec z_{\forall x\varphi}) :\Leftrightarrow \{\vec z\:|\:\:P_{\varphi}^I(\vec z_\varphi) \text{ for all $x\in M$}\}\indent P_{\exists x\varphi}^I(\vec z_{\exists x\varphi}) :\Leftrightarrow \{\vec z\:|\:\:P_{\varphi}^I(\vec z_\varphi) \text{ for some $x\in M$}\}
	\end{matrix}$$
	and for intuitionistic logic and each world $u$
	$$\begin{matrix}
		P_A^{I_u}(\vec z_A):\Leftrightarrow A^{I_u}(\vec z_A)\\
		P_{\varphi\wedge\psi}^{I_u}(\vec z_{\varphi\wedge\psi}) :\Leftrightarrow P_{\varphi}^{I_u}(\vec z_\varphi)\wedge P_{\psi}^{I_u}(\vec z_\psi)\indent P_{\varphi\vee\psi}^{I_u}(\vec z_{\varphi\vee\psi}) :\Leftrightarrow P_{\varphi}^{I_u}(\vec z_\varphi)\vee P_{\psi}^{I_u}((\vec z_\psi))\\
		P_{\varphi\to\psi}^{I_u}(\vec z_{\varphi\to\psi}) :\Leftrightarrow(\neg P_{\varphi}^{I_w}(\vec z_{\varphi}))\vee P_{\psi}^{I_w}(\vec z_{\psi})\text{ for all $w\geq u$}\\
		P_{\forall x\varphi}^{I_u}(\vec z_{\forall x\varphi}) :\Leftrightarrow P_{\varphi}^{I_w}(\vec z_\varphi) \text{ for all $w\geq u$, $x\in M_w$}\indent P_{\exists x\varphi}^{I_u}(\vec z_{\exists x\varphi}) :\Leftrightarrow P_{\varphi}^{I_u}(\vec z_\varphi) \text{ for some $x\in M_u$}
	\end{matrix}$$
\end{definition}

@@

\begin{example}\label{ex:prop-tree-model}
	This is an example for the model translation from Corollary~\ref{cor:prop-tree-model}. Consider
	$$\mathcal S = \{(A\to B)\to C, (B\to A)\to D, (A\wedge B)\to \bot, (C\vee D)\to E\}$$and $\varphi = \bigwedge S\to E$ which has the Kripke counter-model, depicted below,
	\begin{center}
		\begin{tikzpicture}[squarednode/.style={rectangle, draw, very thick, minimum size=5mm}]
			\node (center) {};
			\node[squarednode]      (right)   [right=of center] {$B, C$};
			\node[squarednode]      (left)    [left=of center] {$A, D$};
			\node[squarednode]      (lower)       	[below=of center] {};
			\draw[-{Stealth[length=3mm, width=3mm]}] (lower.north west) -- (left.south east);
			\draw[-{Stealth[length=3mm, width=3mm]}] (lower.north east) -- (right.south west);			
		\end{tikzpicture}
	\end{center}
	\vspace*{-.3cm}where at each node the true propositions are indicated.
	%	
	There is a corresponding classical counter-model $\mathcal M = (M, I)$ to $\varphi^\circ$ with $M = \{u, u_l, u_r\}$, $A^{I}(l) = D^I(l) = B^I(r) = C^I(r) = 1$ and $0$ else.
	%
	Denoting $\psi_1 := (A\to B)\to C$ and $\psi_2 := (A\to B)\to D$, this corresponds to a transformed counter-model $\mathcal M_T = (M_T, I_T)$ of $\varphi^\circ$ with $M_T = \{\epsilon, \psi_1, \psi_2, \psi_1\psi_2, \psi_2\psi_1\}$ and interpretation as presented below:
	\begin{center}
		\begin{tikzpicture}[squarednode/.style={rectangle, draw, very thick, minimum size=5mm}]
			\node (center) {};
			\node[squarednode]      (right1)   [right=of center] {$B, C$};
			\node[squarednode]      (right2)   [above=of right1] {$B, C$};
			\node[squarednode]      (left1)    [left=of center] {$A, D$};
			\node[squarednode]      (left2)    [above=of left1] {$A, D$};
			\node[squarednode]      (lower)       	[below=of center] {};
			\draw[-{Stealth[length=3mm, width=3mm]}] (lower.north west) -- (left1.south east) node [midway, below left] {$f_{\psi_1}$};
			\draw[-{Stealth[length=3mm, width=3mm]}] (lower.north east) -- (right1.south west) node [midway, below right] {$f_{\psi_2}$};
			\draw[-{Stealth[length=3mm, width=3mm]}] (left1.north) -- (left2.south) node [midway, right] {$f_{\psi_2}$};
			\draw[-{Stealth[length=3mm, width=3mm]}] (right1.north) -- (right2.south) node [midway, left] {$f_{\psi_1}$};
			\draw[-{Stealth[length=3mm, width=3mm]}] (left1.south west) to [out=225, in=135, loop, looseness=3] node [left] {$f_{\psi_1}$} (left1.north west);		
			\draw[-{Stealth[length=3mm, width=3mm]}] (right1.south east) to [out=315, in=45, loop, looseness=3] node [right] {$f_{\psi_2}$} (right1.north east);
			\draw[-{Stealth[length=3mm, width=3mm]}] (left2.north west) to [out=135, in=45, loop, looseness=3] node [midway, above] {$f_{\psi_1}, f_{\psi_2}$} (left2.north east);
			\draw[-{Stealth[length=3mm, width=3mm]}] (right2.north west) to [out=135, in=45, loop, looseness=3] node [midway, above] {$f_{\psi_1}, f_{\psi_2}$} (right2.north east);		
		\end{tikzpicture}
	\end{center}
\end{example}

\begin{lemma}\label{lemma:QBF}
	$\varphi$ is not intuitionistically valid iff $\varphi^Q$ is a satisfiable QBF.
\end{lemma}
\begin{proof}
	Suppose $\varphi$ is not intuitionistically valid i.e. it $\bigwedge\mathcal S\to P$ has an intuitionistic counter-model. Then by the precious section there exists a classical counter-model $\mathcal M$ for $\bigwedge\mathcal S^\#\to P^\epsilon$.
	For each atom $A$ interpret $A^0$ such as $\mathcal M$ interprets $A^\epsilon$. Suppose we are given interpretations of all atoms $A^i$ for $i < n$ and a sequence with no repetitions $\psi_1\dots\psi_{n-1}$ over $\mathcal F_\to$ such that $\psi_i$ is exactly the $\psi\in\mathcal F_\to$ for which $F_{\psi}^i$ is true and $A^i$ is interpreted as $A^{\psi_1\dots\psi_i}$ is in $\mathcal M$.

	Let the $F^{n}_\psi$ be arbitrarily interpreted (since they are $\forall$-quantified). If not exactly one is interpreted as true then valid$(n)$ fails, i.e. have $\psi_n\in\mathcal F_\to$ the only element with $F^n_{\psi_n} = 1$. If $F^i_{\psi_n} = 1$ for some $i < n$ then valid$(n)$ also fails, so we may assume that $\psi_1\dots\psi_n$ is a sequence with no repetitions. Interpret the atoms $A^n$ as $\mathcal M$ interprets $A^{\psi_1\dots\psi_n}$. Continue this construction until $n  = |\mathcal F_\to|$. Then from $\mathcal M$ being a counter-example to $\mathcal S^\#\to P$ it directly follows that this interpretation satisfies $\varphi^Q$.

	On the other hand suppose $\varphi^Q$ is satisfiable. We construct a counter-example to $\varphi^\#$.
	Again we proceed iteratively. Interpret $A^\epsilon$ such as $A^0$ is interpreted in some satisfying interpretation of $\varphi^Q$. Suppose we are given a sequence $\psi_1\dots \psi_{n-1}$ such that for $i<n$ having $A^i = A^{\psi_0\dots\psi_{n-1}}$ is part of a satisfying interpretation of $\varphi^Q$ in which $F^i_\psi$ is chosen true iff $\psi = \psi_i$. Let $\psi_n\in\mathcal F_\to$. Consider some interpretation of the $A^n$ that are part of a satisfying assignment where $F^n_\psi$ is true iff $\psi = \psi_n$ and all variables quantified above are chosen as before. Have $A^{\psi_0\dots\psi_n} = A^n$ for each propositional variable $A$. From the definitions it directly follows that construction yields a counter-model for $\varphi^\#$.
\end{proof}

>>>>

Suppose $\mathcal M, v\models\chi$. Then, for all $m\in M$ we have that $\mathcal M, v[m/a]\models\varphi[a/x]$ and therefore $\mathcal M_{\chi}, v[m/a]\models\varphi[a/x]^S_{Z\cup\{a\}}$ and thus $\mathcal M_{\chi}, v\models \forall x\varphi([a/x]^S_{Z\cup\{a\}}[x/a])$, i.e. $\mathcal M_\chi,v\models \chi^S_Z$. On the other hand suppose $\mathcal M, v\not\models\chi$. Then there exists $m\in M$ such that $\mathcal M, v[m/a]\not\models\varphi[a/x]$ and therefore $\mathcal M_\chi, v[m/a]\not\models\varphi[a/x]^H_{Z\cup\{a\}}$ and so by definition $\mathcal M_\chi, v\not\models\varphi[s(z_1,\dots z_n)/x]^H_Z$, i.e. $\mathcal M_\chi, v\not\models(\forall x\varphi)^H_Z$.

@@

\begin{definition}\label{def:transf-structure}
	For every classical and intuitionistic structure $\mathcal M$ define a structure $\mathcal S(\mathcal M,\varphi)$ that agrees with $\mathcal M$ on everything except the interpretation(s) of atoms of the form $P_\psi$. By slight abuse of notation we denote with $\vec z_\psi$ elements of the domain instead of variables. In the classical case, we set
	$$\begin{matrix}
		P_A^I(\vec z_A):\Leftrightarrow A^I(\vec z_A)\\
		P_{\varphi\wedge\psi}^I(\vec z_{\varphi\wedge\psi}) :\Leftrightarrow P_{\varphi}^I(\vec z_\varphi)\wedge P_{\psi}^I(\vec z_\psi)\indent P_{\varphi\vee\psi}^I(\vec z_{\varphi\vee\psi}) :\Leftrightarrow P_{\varphi}^I(\vec z_\varphi)\vee P_{\psi}^I((\vec z_\psi))\\P_{\varphi\to\psi}^I(\vec z_{\varphi\to\psi}) :\Leftrightarrow (\neg P_{\varphi}^I(\vec z_{\varphi}))\vee P_{\psi}^I(\vec z_{\psi})\\P_{\forall x\varphi}^I(\vec z_{\forall x\varphi}) :\Leftrightarrow \{\vec z\:|\:\:P_{\varphi}^I(\vec z_\varphi) \text{ for all $x\in M$}\}\indent P_{\exists x\varphi}^I(\vec z_{\exists x\varphi}) :\Leftrightarrow \{\vec z\:|\:\:P_{\varphi}^I(\vec z_\varphi) \text{ for some $x\in M$}\},
	\end{matrix}$$
	and for intuitionistic logic and each world $u$, we set
	$$\begin{matrix}
		P_A^{I_u}(\vec z_A):\Leftrightarrow A^{I_u}(\vec z_A)\\
		P_{\varphi\wedge\psi}^{I_u}(\vec z_{\varphi\wedge\psi}) :\Leftrightarrow P_{\varphi}^{I_u}(\vec z_\varphi)\wedge P_{\psi}^{I_u}(\vec z_\psi)\indent P_{\varphi\vee\psi}^{I_u}(\vec z_{\varphi\vee\psi}) :\Leftrightarrow P_{\varphi}^{I_u}(\vec z_\varphi)\vee P_{\psi}^{I_u}((\vec z_\psi))\\
		P_{\varphi\to\psi}^{I_u}(\vec z_{\varphi\to\psi}) :\Leftrightarrow(\neg P_{\varphi}^{I_w}(\vec z_{\varphi}))\vee P_{\psi}^{I_w}(\vec z_{\psi})\text{ for all $w\geq u$}\\
		P_{\forall x\varphi}^{I_u}(\vec z_{\forall x\varphi}) :\Leftrightarrow P_{\varphi}^{I_w}(\vec z_\varphi) \text{ for all $w\geq u$, $x\in M_w$}\indent P_{\exists x\varphi}^{I_u}(\vec z_{\exists x\varphi}) :\Leftrightarrow P_{\varphi}^{I_u}(\vec z_\varphi) \text{ for some $x\in M_u$}.
	\end{matrix}$$
\end{definition}

@@

\begin{example}\label{ex:prop-tree-model}
	This is an example for the model translation from Corollary~\ref{cor:prop-tree-model}. Consider
	$$\mathcal S = \{(A\to B)\to C, (B\to A)\to D, (A\wedge B)\to \bot, (C\vee D)\to E\}$$and $\varphi = \bigwedge S\to E$ which has the Kripke counter-model, depicted below,
	\begin{center}
		\begin{tikzpicture}[squarednode/.style={rectangle, draw, very thick, minimum size=5mm}]
			\node (center) {};
			\node[squarednode]      (right)   [right=of center] {$B, C$};
			\node[squarednode]      (left)    [left=of center] {$A, D$};
			\node[squarednode]      (lower)       	[below=of center] {};
			\draw[-{Stealth[length=3mm, width=3mm]}] (lower.north west) -- (left.south east);
			\draw[-{Stealth[length=3mm, width=3mm]}] (lower.north east) -- (right.south west);			
		\end{tikzpicture}
	\end{center}
	\vspace*{-.3cm}where at each node the true propositions are indicated.
	%	
	There is a corresponding classical counter-model $\mathcal M = (M, I)$ to $\varphi^\circ$ with $M = \{u, u_l, u_r\}$, $A^{I}(l) = D^I(l) = B^I(r) = C^I(r) = 1$ and $0$ else.
	%
	Denoting $\psi_1 := (A\to B)\to C$ and $\psi_2 := (B\to A)\to D$, this corresponds to a transformed counter-model $\mathcal M_T = (M_T, I_T)$ of $\varphi^\circ$ with $M_T = \{\epsilon, \psi_1, \psi_2, \psi_1\psi_2, \psi_2\psi_1\}$ and interpretation as presented below:
	\begin{center}
		\begin{tikzpicture}[squarednode/.style={rectangle, draw, very thick, minimum size=5mm}]
			\node (center) {};
			\node[squarednode]      (right1)   [right=of center] {$B, C$};
			\node[squarednode]      (right2)   [above=of right1] {$B, C$};
			\node[squarednode]      (left1)    [left=of center] {$A, D$};
			\node[squarednode]      (left2)    [above=of left1] {$A, D$};
			\node[squarednode]      (lower)       	[below=of center] {};
			\draw[-{Stealth[length=3mm, width=3mm]}] (lower.north west) -- (left1.south east) node [midway, below left] {$f_{\psi_1}$};
			\draw[-{Stealth[length=3mm, width=3mm]}] (lower.north east) -- (right1.south west) node [midway, below right] {$f_{\psi_2}$};
			\draw[-{Stealth[length=3mm, width=3mm]}] (left1.north) -- (left2.south) node [midway, right] {$f_{\psi_2}$};
			\draw[-{Stealth[length=3mm, width=3mm]}] (right1.north) -- (right2.south) node [midway, left] {$f_{\psi_1}$};
			\draw[-{Stealth[length=3mm, width=3mm]}] (left1.south west) to [out=225, in=135, loop, looseness=3] node [left] {$f_{\psi_1}$} (left1.north west);		
			\draw[-{Stealth[length=3mm, width=3mm]}] (right1.south east) to [out=315, in=45, loop, looseness=3] node [right] {$f_{\psi_2}$} (right1.north east);
			\draw[-{Stealth[length=3mm, width=3mm]}] (left2.north west) to [out=135, in=45, loop, looseness=3] node [midway, above] {$f_{\psi_1}, f_{\psi_2}$} (left2.north east);
			\draw[-{Stealth[length=3mm, width=3mm]}] (right2.north west) to [out=135, in=45, loop, looseness=3] node [midway, above] {$f_{\psi_1}, f_{\psi_2}$} (right2.north east);		
		\end{tikzpicture}
	\end{center}
\end{example}

\begin{lemma}\label{lemma:QBF}
	$\varphi$ is not intuitionistically valid iff $\varphi^Q$ is a satisfiable QBF.
\end{lemma}
\begin{proof}
Suppose $\varphi$ is not intuitionistically valid.
Then, $\bigwedge\mathcal S\to P$ has an intuitionistic counter-model. Then by the previous section there exists a classical counter-model $\mathcal M$ for $\bigwedge\mathcal S^\#\to P^\epsilon$.
We now define a QBF model iteratively.
For each atom $A$ interpret $A^0$ such as $\mathcal M$ interprets $A^\epsilon$. Suppose we are given interpretations of all atoms $A^i$ for $i < n$ and a sequence with no repetitions $\psi_1\dots\psi_{n-1}$ over $\mathcal F_\to$ such that $\psi_i$ is exactly the $\psi\in\mathcal F_\to$ for which $F_{\psi}^i$ is true and $A^i$ is interpreted as $A^{\psi_1\dots\psi_i}$  in $\mathcal M$.
Let the $F^{n}_\psi$ be arbitrarily interpreted (since they are $\forall$-quantified). If not exactly one of the $F^{n}_\psi$ is interpreted as true, then valid$(n)$ fails and the remaining propositions can be chosen arbitrarily. If $F^i_{\psi_n} = 1$ for some $i < n$, then valid$(n)$ also fails and the remaining propositions can be chosen arbitrarily.
So we may assume that $\psi_1\dots\psi_n$ is a sequence with no repetitions. 
Interpret the atoms $A^n$ as $\mathcal M$ interprets $A^{\psi_1\dots\psi_n}$. 
Continue this construction until $n  = |\mathcal F_\to|$. Then from $\mathcal M$ being a counter-example to $\mathcal S^\#\to P$ it directly follows that this interpretation satisfies $\varphi^Q$.

	On the other hand suppose $\varphi^Q$ is satisfiable. We construct a counter-example to $\varphi^\#$.
	Again we proceed iteratively. Interpret $A^\epsilon$ such as $A^0$ is interpreted in some satisfying interpretation of $\varphi^Q$. Suppose we are given a sequence $\psi_1\dots \psi_{n-1}$ such that for $i<n$ having $A^i = A^{\psi_0\dots\psi_{n-1}}$ is part of a satisfying interpretation of $\varphi^Q$, in which $F^i_\psi$ is chosen true iff $\psi = \psi_i$. Let $\psi_n\in\mathcal F_\to$. Consider some interpretation of the $A^n$ that are part of a satisfying assignment where $F^n_\psi$ is true iff $\psi = \psi_n$ and all variables quantified above are chosen as before. Have $A^{\psi_0\dots\psi_n} = A^n$ for each propositional variable $A$. From the definitions it directly follows that construction yields a counter-model for $\varphi^\#$.
\end{proof}
